---
title: "NPS Zooplankton Exploration"
author: "Matt Brousil"
date: "August 19, 2019"
output: html_document
---
```{r, include=FALSE}
library(tidyverse)
library(readxl)
library(visdat)
library(vegan)
library(janitor)
```

### Load up the data

```{r}
zoop <- read_excel("../data/NCCN_Zoops_Combined_For_FCA_May_2019.xlsx")
```

There's a row where "Collotheca pelagica" is pasted across three columns
It is treated as an NA and Taxonid is retained as numeric
Admittedly I don't have a full grasp of the last three cols, but I'll make
the record match with the others of that species nearby:

```{r}
zoop[1487, "Taxa"] <- "ROT"
zoop[1487, "Taxonid"] <- 5
```

### Quick view of the whole dataset:
```{r}
# Data types of the data frame
vis_dat(zoop)
# Missing data
vis_miss(zoop)
```

Three additional records where Taxa and Taxonid aren't complete. They make sense:
```{r}
zoop %>% filter(is.na(Taxa) | is.na(Taxonid))
```

### Genus-species by year:
```{r}
zoop_lake_year_gensp <- zoop %>%
  filter(!is.na(Taxa),
         !is.na(Taxonid)) %>%
  ggplot() +
  geom_bar(aes(x = Year, y = Density, fill = GenSp),
           stat = "identity", position = "fill") +
  #geom_point(aes(x = Year, y = Density, color = Taxa)) +
  facet_wrap(. ~ Lake) +
  theme(axis.text.x = element_text(angle = 69, hjust = 1),
        legend.position = "bottom") +
  ggtitle("Relative densities of zooplankton species by lake and year")

zoop_lake_year_gensp + theme(legend.position = "none")

ggsave(filename = "../figures/paneled_by_lake/zoop_year_dens_gensp.png",
       plot = zoop_lake_year_gensp, device = "png", width = 12, height = 12,
       units = "in")
```

### Taxa by year:
```{r}
zoop_lake_year_taxa <- zoop %>%
  filter(!is.na(Taxa),
         !is.na(Taxonid)) %>%
  ggplot() +
  geom_bar(aes(x = Year, y = Density, fill = Taxa),
           stat = "identity", position = "fill") +
  #geom_point(aes(x = Year, y = Density, color = Taxa)) +
  facet_wrap(. ~ Lake) +
  theme(axis.text.x = element_text(angle = 69, hjust = 1),
        legend.position = "right") +
  ggtitle("Relative densities of zooplankton taxa by lake and year")

zoop_lake_year_taxa

ggsave(filename = "../figures/paneled_by_lake/zoop_year_dens_taxa.png",
       plot = zoop_lake_year_taxa, device = "png", width = 12, height = 12,
       units = "in")
```

## Try out some multivariate stuff

### Communities across all years at the species level. Doesn't work:
```{r}
# Remove sp making up < 5%
filtered_zoop <- zoop %>%
  clean_names() %>%
  filter(!is.na(taxa)) %>%
  group_by(year, park, lake, taxa, gen_sp) %>%
  summarise(sp_sum = sum(density)) %>%
  group_by(year, park, lake) %>%
  mutate(total_sum = sum(sp_sum, na.rm = T)) %>%
  ungroup() %>%
  mutate(prop = sp_sum / total_sum) %>%
  filter(prop >= .05)

zoop_mat <- filtered_zoop %>%
  spread(key = gen_sp, value = sp_sum, fill = 0) %>%
  data.frame() %>%
  clean_names() %>%
  select(bosmina_longirostris:synchaeta_sp)

metaMDS(comm = zoop_mat, k = 2, try = 20)
```

### Communities at higher taxa level. Still doesn't work:
```{r}
zoop_mat_taxa <-  filtered_zoop %>%
  filter(!is.na(taxa)) %>%
  group_by(year, park, lake, taxa) %>%
  summarise(taxa_sum = sum(sp_sum)) %>%
  # Lines below removed bc no point in removing GenSp or Taxa twice? Already
  # done for filtered_zoop
  # mutate(prop = taxa_sum / sum(taxa_sum, na.rm = T)) %>%
  # filter(prop >= .05) %>%
  # select(-prop) %>%
  spread(key = taxa, value = taxa_sum, fill = 0) %>%
  data.frame() %>%
  select(CLAD:ROT)

metaMDS(comm = zoop_mat_taxa, k = 2, try = 20)
```

Overview of the taxa-level dataset:
```{r}
summary(filtered_zoop %>%
  filter(!is.na(taxa)) %>%
  group_by(year, park, lake, taxa) %>%
  summarise(taxa_sum = sum(sp_sum)) %>%
  mutate(prop = taxa_sum / sum(taxa_sum, na.rm = T)) %>%
  filter(prop >= .05) %>%
  spread(key = taxa, value = taxa_sum, fill = 0) %>%
  data.frame() %>%
  select(CLAD:ROT))

```


### NMDS for each lake separately, with zoops at species level: 
```{r}

spread_species <- filtered_zoop %>%
  spread(key = gen_sp, value = sp_sum, fill = 0)

map(.x = unique(filtered_zoop$lake),
    ~ metaMDS(comm = filter(spread_species, lake == .) %>%
                data.frame() %>%
                clean_names() %>%
                select(-c(year:prop)),
              k = 2, try = 20))

```

### NMDS for each lake separately, with zoops at higher taxa level: 
```{r}

spread_taxa <- filter(filtered_zoop, !is.na(taxa)) %>%
                group_by(year, park, lake, taxa) %>%
                summarise(taxa_sum = sum(sp_sum)) %>%
                mutate(prop = taxa_sum / sum(taxa_sum, na.rm = T)) %>%
                filter(prop >= .05) %>%
                spread(key = taxa, value = taxa_sum, fill = 0)

map(.x = unique(filtered_zoop$lake),
    ~ metaMDS(comm = filter(spread_taxa, lake == .) %>%
                data.frame() %>%
                select(CLAD:ROT),
              k = 2, try = 20))

```

## CCA of environmental vars with zoops

## CCA of watershed chars and multivariate lake vars

## Zoop aggregation (ratios)




















