---
title: "Alternative zooplankton exploration"
author: "Matt Brousil"
date: "August 21, 2019"
output: html_document
---
```{r, include=FALSE}
library(tidyverse)
library(readxl)
library(visdat)
library(vegan)
library(janitor)
```

### Load up the data

```{r}
zoop <- read_excel("../data/NCCN_Zoops_Combined_For_FCA_May_2019.xlsx")
```

There's a row where "Collotheca pelagica" is pasted across three columns
It is treated as an NA and Taxonid is retained as numeric
Admittedly I don't have a full grasp of the last three cols, but I'll make
the record match with the others of that species nearby:

```{r}
zoop[1487, "Taxa"] <- "ROT"
zoop[1487, "Taxonid"] <- 5
```

### Quick view of the whole dataset:
```{r}
# Data types of the data frame
vis_dat(zoop)
# Missing data
vis_miss(zoop)
```

Three additional records where Taxa and Taxonid aren't complete. They make sense, but should mosquito pupae be retained?
```{r}
zoop %>% filter(is.na(Taxa) | is.na(Taxonid))
```

```{r}
taxa_data <- zoop %>%
     filter(!is.na(Taxa)) %>%
 group_by(Year, Park, Lake, Taxa) %>%
   summarise(taxa_sum = sum(Density)) %>%
 spread(key = Taxa, value = taxa_sum, fill = 0) %>%
     data.frame() %>%
     select(CLAD:ROT)

head(taxa_data)
```

NMDS:
```{r}
taxa_nmds <- metaMDS(comm = taxa_data, k = 2, try = 20)

ordiplot(ord = taxa_nmds, display = "sites")
```