---
title: "Zooplankton and Environmental Data Analysis"
author: "Matt Brousil"
date: "August 28, 2019"
output: html_document
---

```{r echo = FALSE, results = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(readxl)
library(visdat)
library(vegan)
library(janitor)
```

## 1. Zooplankton communities ~ environmental variables

### 1b. Zooplankton CCA  

**Species level**  
```{r, echo = FALSE, results = 'hide', fig.show='hide', warning=FALSE}
# Load data
bigjoin <- readRDS(file = "../data/analysis_outputs/bigjoin.rds")
zoop <- read_excel("../data/NCCN_Zoops_Combined_For_FCA_May_2019.xlsx")

#  Prep data
zoop[1487, "Taxa"] <- "ROT"
zoop[1487, "Taxonid"] <- 5

# Name matching table
match_sites <- readRDS(file = file.path("..",
                                        "data",
                                        "name_site_matches.rds"))

# Crosswalk common names for sites to site_codes
zoop <- left_join(x = zoop, y = match_sites,
                               by = c("Lake" = "old_name"))

# Grab yearly environmental data...quick check

bigjoin %>%
  filter(variable %in% c("secchi_value_m", "AirTemp", "BotTemp",
                         "MidTemp", "SurfTemp", "DO_top2m")) %>%
  select(park_code, site_code, event_year, start_date, ice_out_doy) %>%
  unique() %>%
  count(park_code, site_code, event_year, start_date) %>%
  arrange(desc(n))

yearly_bigjoin <- bigjoin %>%
  filter(variable %in% c("secchi_value_m", "AirTemp", "BotTemp",
                         "MidTemp", "SurfTemp", "DO_top2m")) %>%
  select(park_code, site_code, event_year, variable, value) %>%
  unique()

head(yearly_bigjoin)  
head(zoop)

# There's 114 rows in the yearly_bigjoin that don't match up with zoop data
zoop_env <- inner_join(x = zoop, y = yearly_bigjoin,
          by = c("site_code",
                 "Year" = "event_year")) %>%
  spread(key = variable, value = value)

# Double check that there's not multiple measures/year (Spoiler: There are.)
zoop_env %>%
  count(park_code, site_code, Year, GenSp) %>%
  arrange(desc(n))

# Deadwood has multiple samples per zoop species per year? I'm not clear
# as to why this is. Perhaps I'm missing something. Going to sum them up
# for now.
zoop_env %>%
  filter(park_code == "MORA", site_code == "LW32", Year == "2018")

zoop_sum <- zoop %>%
  group_by(Park, Year, Lake, site_code, Code, Taxa, Taxonid, GenSp) %>%
  summarise(density = sum(Density))
  
zoop_env_sum <- inner_join(x = zoop_sum, y = yearly_bigjoin,
                       by = c("site_code",
                              "Year" = "event_year")) %>%
  spread(key = variable, value = value)

# Looks good:
zoop_env_sum %>%
  count(park_code, site_code, Year, GenSp) %>%
  arrange(desc(n))

# Variables in the dataset currently:
names(zoop_env_sum)

zoop_env_sp_no_na_spread <- zoop_env_sum %>%
  filter_at(vars(AirTemp:SurfTemp), all_vars(!is.na(.))) %>%
  spread(key = GenSp, value = density)

vis_dat(zoop_env_sp_no_na_spread)

# Not sure this is the way to go...Looking at the absence of data in this df
# it's clear that the species are fairly unique to specific places? Would still
# fill in the NAs with 0s, but still there's a lot of zero data. 

cca_comm_data_sp <- zoop_env_sp_no_na_spread %>%
  clean_names() %>%
  data.frame() %>%
  select(alona_costata:trichotria_tetractis) %>%
  # Replace NAs with 0s
  map_df(.x = .,  ~ replace_na(data = ., replace = 0))

cca_env_data_sp <- zoop_env_sp_no_na_spread %>%
  data.frame() %>%
  select(AirTemp:SurfTemp)

zoop_cca_sp <- cca(X = as.matrix(cca_comm_data_sp),
                Y = as.matrix(cca_env_data_sp))
```


CCA plot for Zoop sp density ~ environmental data
```{r}
ordiplot(zoop_cca_sp)
```

**Taxa level**
```{r, echo = FALSE, results = 'hide', fig.show='hide'}
# Let's try taxa level too. Going to backup a couple steps...

zoop_sum_taxa <- zoop %>%
  # Dropping Code b/c it appears redundant with Taxa
  group_by(Park, Year, Lake, site_code, Taxa, Taxonid) %>%
  summarise(density = sum(Density))

zoop_env_sum_taxa <- inner_join(x = zoop_sum_taxa, y = yearly_bigjoin,
                           by = c("site_code",
                                  "Year" = "event_year")) %>%
  spread(key = variable, value = value)

# Looks good:
zoop_env_sum_taxa %>%
  count(park_code, site_code, Year, Taxa) %>%
  arrange(desc(n))

# Variables in the dataset currently:
names(zoop_env_sum_taxa)

# CCA will only run if there are no NA values...I don't think it makes sense
# to turn NA values for the env data into 0s. So, are there rows where all
# predictors aren't NA?
zoop_env_sum_taxa %>%
  filter(!is.na(Taxa)) %>%
  spread(key = Taxa, value = density)

# Yes.
vis_dat(zoop_env_sum_taxa %>%
  filter(!is.na(Taxa)) %>%
  spread(key = Taxa, value = density))

# **How to account (at all?) account for the fact that not all lakes have
# identical timelines of zoop data, and some that do will probably have gaps in
# their environmental data?**
zoop_env_sum_taxa %>%
  filter_at(vars(AirTemp:SurfTemp), all_vars(!is.na(.))) %>%
  # What space and time spread to they cover?
  count(Park, Lake, Year) %>%
  ggplot() +
  geom_point(aes(x = Year, y = Lake)) +
  facet_grid(rows = vars(Park))

# Create a version of the dataset that has only data with no predictor NAs
zoop_env_taxa_no_na_spread <- zoop_env_sum_taxa %>%
  filter(!is.na(Taxa)) %>%
  select(-Taxonid) %>%
  filter_at(vars(AirTemp:SurfTemp), all_vars(!is.na(.))) %>%
  spread(key = Taxa, value = density)

cca_comm_data_taxa <- zoop_env_taxa_no_na_spread %>%
  data.frame() %>%
  select(CLAD:ROT) %>%
  # Replace NAs with 0s
  map_df(.x = .,  ~ replace_na(data = ., replace = 0))

cca_env_data_taxa<- zoop_env_taxa_no_na_spread %>%
  data.frame() %>%
  select(AirTemp:SurfTemp)

zoop_cca_taxa <- cca(X = as.matrix(cca_comm_data_taxa),
                Y = as.matrix(cca_env_data_taxa))
```

CCA plot for Zoop taxa density ~ environmental data
```{r}
ordiplot(zoop_cca_taxa)
```

### 1c. Zooplankton DBRDA (`capscale()`)
```{r}
library(tidyverse)
library(readxl)
library(visdat)
library(vegan)
library(janitor)
library(psych)
library(corrplot)
library(ggvegan)

bigjoin <- readRDS(file = "../data/analysis_outputs/bigjoin.rds")
zoop <- read_excel("../data/NCCN_Zoops_Combined_For_FCA_May_2019.xlsx")

#  Prep data
zoop[1487, "Taxa"] <- "ROT"
zoop[1487, "Taxonid"] <- 5

# Notes on the zoops: Going to avoid species in this analysis because many zeroes.
# Also, there appear to be some synonyms.
# If synonyms figured out, could slim things down by selecting >= 5% contribs

# Name matching table
match_sites <- readRDS(file = file.path("..",
                                        "data",
                                        "name_site_matches.rds"))

# Crosswalk common names for sites to site_codes
zoop <- left_join(x = zoop, y = match_sites,
                  by = c("Lake" = "old_name")) %>%
  clean_names()

# Turn density measures into proportions of each taxon per sample. Filter out
# those providing <5%
zoop_prop_taxa <- zoop %>%
  # Dropping Code for now b/c unclear what it means. Code != Taxa though.
  # Also dropping taxonid
  group_by(park, year, lake, site_code, taxa) %>%
  summarise(density = sum(density)) %>%
  group_by(park, year, site_code) %>%
  mutate(sum_density = sum(density, na.rm = T)) %>%
  ungroup() %>%
  mutate(prop_density = density / sum_density) %>%
  filter(prop_density >= .05)

yearly_bigjoin <- bigjoin %>%
  filter(variable %in% c("secchi_value_m", "AirTemp", "BotTemp",
                         "MidTemp", "SurfTemp", "DO_top2m")) %>%
  select(park_code, site_code, event_year, variable, value) %>%
  unique()

zoop_env_prop_taxa <- inner_join(x = zoop_prop_taxa, y = yearly_bigjoin,
                                by = c("site_code",
                                       "year" = "event_year")) %>%
  spread(key = variable, value = value)

# Looks good:
zoop_env_prop_taxa %>%
  count(park_code, site_code, year, taxa) %>%
  arrange(desc(n))

# Variables in the dataset currently:
names(zoop_env_prop_taxa)

# Now look at env variable correlations, but first check for and remove NAs
vis_dat(x = zoop_env_prop_taxa)

zoop_env_prop_taxa_full <- na.omit(zoop_env_prop_taxa)

corrplot(cor(select(zoop_env_prop_taxa_full, AirTemp:SurfTemp)), type = "upper", 
         tl.col = "black", tl.srt = 45)

corr.test(x = select(zoop_env_prop_taxa_full, AirTemp:SurfTemp))

# How about keep SurfTemp, drop other temps.
corrplot(cor(select(zoop_env_prop_taxa_full, AirTemp, DO_top2m, secchi_value_m)), type = "upper", 
         tl.col = "black", tl.srt = 45)

corr.test(x = select(zoop_env_prop_taxa_full, AirTemp, DO_top2m, secchi_value_m))

# If I'm understanding the output correctly, DO and secchi are correlated significantly, but the correlation is weak. I keep it for now.

zoop_env_prop_taxa_spread <- zoop_env_prop_taxa_full %>%
  select(-density, -sum_density) %>%
  spread(key = taxa, value = prop_density) %>%
  # Replace NAs with 0s
  map_df(.x = .,  ~ replace_na(data = ., replace = 0))

capscale_taxa_data <- zoop_env_prop_taxa_spread %>%
  select(CLAD:ROT)

capscale_env_data <- zoop_env_prop_taxa_spread %>%
  select(DO_top2m, secchi_value_m, SurfTemp)

```

Take a quick detour to log transform the data. Michael suggested that there might
be an issue with imbalance in the axes due to different scales of the data. An issue
with this is that the DO_top2m column has a negative value (-2.919)...Is this allowed?
**Update**: I tried this, but get a subscript out of bounds error when I run the
capscale and it additionally says "results may be meaningless because data have
negative entries in method "bray""
```{r eval=FALSE}
# Section is included here but not evaluated

capscale_taxa_data <- map_df(.x = capscale_taxa_data,
                             .f = ~ log10(.x + 0.01))

capscale_env_data <- map_df(.x = capscale_env_data,
                            .f = ~ log10(.x + 0.01))

# Remove the NA
na_row <- which(is.na(capscale_env_data), arr.ind = TRUE)[, 1]

capscale_env_data <- capscale_env_data[-na_row, ]
capscale_taxa_data <- capscale_taxa_data[-na_row, ]

```


```{r}
# Do the capscale
zoop_env_capscale <- capscale(formula = capscale_taxa_data ~ DO_top2m +
                                SurfTemp + secchi_value_m,
                              data = capscale_env_data, distance = "bray")

zoop_env_capscale

autoplot(zoop_env_capscale)

```







```{r}
autoplot(metaMDS(comm = capscale_taxa_data, trymax = 100))
stressplot(metaMDS(comm = capscale_taxa_data, trymax = 100))

```



















