---
title: "Analyses with static lake/watershed variables"
author: "Matt Brousil"
date: "7/8/2020"
output:
  html_document:
    toc: true
    toc_float: true
---

## A script that uses PCA and regression tree approaches to classify lakes

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(readxl)
library(factoextra)
library(viridis)
library(vegan)
library(rpart)
library(partykit)
library(rpart.plot)
library(ggrepel)
library(lubridate)
library(ggpubr)

```

```{r}
physio <- read_excel(
  path = file.path("..",
                   "data",
                   "analysis_outputs",
                   "study-site-tables.xlsx"))

# Grab bigjoin for land cover
bigjoin <- readRDS(file = "../data/analysis_outputs/bigjoin.rds")


bigjoin_subset <- bigjoin %>%
  select(park_code, site_code, solar_jas, SWE_May, flush_index_noSWE,
         forest:snowice) %>%
  unique() %>%
  group_by(park_code, site_code) %>%
  summarize_all(.funs = ~ mean(., na.rm = TRUE)) %>%
  left_join(x = ., y = select(bigjoin, site_code, Aspect) %>% unique(),
            by = c("site_code"))

# Name matching table
match_sites <- readRDS(file = file.path("..",
                                        "data",
                                        "name_site_matches.rds"))
```


Select the static variables that go into the analyses
```{r}
# Select continuous variables
physio_subset <- physio %>%
  select(Park_code, Lake, Elevation_m, Watershed_area_to_lake_area_ratio,
         Surface_area_ha, Watershed_area_ha, Depth_mean_m, Volume_m3,
         BlueLineOutlet) %>%
  left_join(x = ., y = match_sites,
            by = c("Lake" = "old_name")) %>%
  select(Park_code, site_code, everything(), -Lake) %>%
  full_join(x = ., y = bigjoin_subset, by = c("Park_code" = "park_code",
                                              "site_code"))

```



### PCA

Scale a numeric subset of the data
```{r}
physio_scale <- physio_subset %>%
  select(Elevation_m:snowice) %>%
  mutate_if(.predicate = is.numeric,
            .funs = ~ scale(.)) %>%
  cbind(Park_code = physio_subset$Park_code, .)

row.names(physio_scale) <- paste0(physio_subset$Park_code,
                                  "_", physio_subset$site_code)
```


#### Run PCA for numeric variables
```{r}
physio_pca <- rda(na.omit(select_if(.tbl = physio_scale,
                                    .predicate = is.numeric)))
```

#### PCA Results:
```{r}
physio_pca
```

```{r, echo=FALSE}
biplot(physio_pca,
       display = c("sites", 
                   "species"),
       type = c("text",
                "points"))
```

Add hulls (currently for park)  
```{r}
# https://rpubs.com/brouwern/veganpca#:~:text=PCA%20(Principal%20Components%20Analysis)%20is,is%20also%20developing%20biplot%20tools).
biplot(physio_pca,
       display = c("sites", 
                   "species"),
       type = c("text",
                "points"))
ordihull(physio_pca,
         group = physio_scale$Park_code)
```


#### Run PCA outside of vegan for another plotting option
```{r}
physio_pca <- prcomp(x = select_if(.tbl = physio_scale,
                                   .predicate = is.numeric),
                     scale. = FALSE)
```

#### PCA Results:
```{r}
physio_pca
```

Scree plot
```{r}
fviz_eig(physio_pca)
```

Bioplot & variable contribution
```{r, fig.show="hold", out.width="50%"}


fviz_pca_biplot(X = physio_pca, col.var = viridis(n = 5)[2],
                col.ind = plasma(n = 5)[4], ggtheme = theme_bw())


fviz_pca_var(X = physio_pca, col.var = "contrib", repel = TRUE,
             gradient.cols = viridis(n = 5, begin = 0, end = 1),
             ggtheme = theme_bw())
```


### Regression tree for chlorophyll ~ static vars

```{r}
mean_chl <- bigjoin %>%
  select(park_code, site_code, event_year, variable, value) %>%
  filter(variable == "Chlorophyll") %>%
  spread(key = variable, value = value) %>%
  group_by(park_code, site_code) %>%
  summarise(mean_chl = mean(Chlorophyll))
```


```{r}
reg_tree_data <- physio_subset %>%
  full_join(x = ., y = mean_chl, by = c("Park_code" = "park_code",
                                        "site_code"))
```

Specify the tree, setting minimum number of obs per node before a split to be 10
```{r}
tree <- rpart(formula = mean_chl ~ Elevation_m +  Watershed_area_to_lake_area_ratio +
                Depth_mean_m + forest + barren + solar_jas + Volume_m3 +
                SWE_May + flush_index_noSWE,
              data = reg_tree_data, minsplit = 10)

plot(x = as.party(tree),
     main = "Chlorophyll",
     terminal_panel = node_boxplot)

plotcp(tree)
ptree <- prune(tree, cp = 0.025);
rpart.plot(ptree, main="Average Chlorophyll")
png("NPS_FCA_chl_tree.png", width=1000, height=800)
```


Based on the cross-validation plot, the ideal tree size is 1...


### Slopes of (top 2m water temp ~ SNOTEL/DAS) ~ static vars

Import Power's data object from his analysis script (v32, ~ line 604) 
```{r}

powers_profile_data <- readRDS(file = "../data/analysis_outputs/sp_dataplot.rds")

powers_profile_data <- powers_profile_data %>%
  separate(data = .,
           col = park_site,
           into = c("park", "site"),
           sep = " ",
           extra = "merge")

plot_wtempSWE_snotel <- ggplot(data = powers_profile_data,
                               aes(x = SWE_May_snotel, y = value, group = "black",
                                   color = (park_code), label = short_code)) +
  geom_line(aes(group = short_code), alpha = 0.5)+
  geom_text_repel(data = filter(powers_profile_data, snowyrlo_snotel == 1),
                  size = 2.5, segment.color = "black", alpha = 1,
                  segment.size = 0.2, box.padding = 0.1) + 
  xlab("May SWE (cm), SNOTEL")+
  ylab(paste0("Water temperature (", intToUtf8(176), "C), top 2m"))+
  theme_bw()+
  theme(legend.position = "none")+
  theme(strip.text.x = element_text(),
        axis.title.x = element_text(vjust = -0.5)) + 
  scale_colour_viridis_d(end = 0.85) +
  xlim(x = c(0, 325))

plot_wtempSWE_snodas <- ggplot(powers_profile_data,
                               aes(x = SWE_May_snodas, y = value,
                                   group = "black", color = (park_code),
                                   label = short_code)) +
  geom_line(aes(group = short_code), alpha = 0.5)+
  geom_text_repel(data = filter(powers_profile_data, snowyrlo_snotel == 1),
                  size = 2.5, segment.color = "black", alpha = 1,
                  segment.size = 0.2, box.padding = 0.1) + 
  xlab("May SWE (cm), SNODAS") +
  ylab(paste0("Water temperature (", intToUtf8(176), "C), top 2m"))+
  theme_bw() +
  theme(legend.position = "none") +
  theme(strip.text.x = element_text(),
        axis.title.x = element_text(vjust = -0.5)) + 
  scale_colour_viridis_d(end = 0.85) +
  xlim(x = c(0, 325))

```

```{r}

lake_list <- split(powers_profile_data, f = powers_profile_data$short_code)

snodas_models <- map_df(.x = lake_list,
                        .f = ~ {
                          
                          model_output <- lm(formula = value  ~ SWE_May_snodas,
                                             data = .x)$coefficients[2]
                          
                          output <- tibble(
                            park = unique(.x$park),
                            site = unique(.x$site),
                            slope = model_output)
                          
                          return(output)
                          
                        })

snotel_models <- map_df(.x = lake_list,
                        .f = ~ {
                          
                          model_output <- lm(formula = value  ~ SWE_May_snotel,
                                             data = .x)$coefficients[2]
                          
                          output <- tibble(
                            park = unique(.x$park),
                            site = unique(.x$site),
                            slope = model_output)
                          
                          return(output)
                          
                        })

```

```{r}
# Going to recreate SP's plot but using actual lm() outputs to make sure that
# the regression coefficient values are accurate:
# https://stackoverflow.com/questions/44865508/using-ggplot2-to-plot-an-already-existing-linear-model

snodas_predict <-  map_df(.x = lake_list,
                          .f = ~ {
                            
                            model_output <- lm(formula = value  ~ SWE_May_snodas,
                                               data = .x)
                            
                            output <- cbind(.x,
                                            fit_source = "snodas",
                                            fit = predict(model_output),
                                            slope = round(model_output$coefficients[2],
                                                          digits = 2))
                            
                            return(output)
                            
                          })

snodas_predict_plot <- snodas_predict %>%
  ggplot(data = .,
         aes(x = SWE_May_snodas, y = fit,
             group = "black", color = (park_code))) +
  geom_line(aes(group = short_code), alpha = 0.5)+
  geom_text_repel(data = filter(snodas_predict, snowyrlo_snotel == 0),
                  size = 2.5, segment.color = "black", alpha = 1,
                  segment.size = 0.2, box.padding = 0.1,
                  mapping = aes(label = paste0(short_code, "_", slope))) + 
  xlab("May SWE (cm), SNODAS") +
  ylab(paste0("Water temperature (", intToUtf8(176), "C), top 2m"))+
  theme_bw() +
  theme(legend.position = "none") +
  theme(strip.text.x = element_text(),
        axis.title.x = element_text(vjust = -0.5)) + 
  scale_colour_viridis_d(end = 0.85) +
  xlim(x = c(0, 325))

snotel_predict <-  map_df(.x = lake_list,
                          .f = ~ {
                            
                            model_output <- lm(formula = value  ~ SWE_May_snotel,
                                               data = .x)
                            
                            output <- cbind(.x,
                                            fit_source = "snotel",
                                            fit = predict(model_output),
                                            slope = round(model_output$coefficients[2],
                                                          digits = 2))
                            
                            return(output)
                            
                          })

snotel_predict_plot <- snotel_predict %>%
  ggplot(data = .,
         aes(x = SWE_May_snotel, y = fit,
             group = "black", color = (park_code))) +
  geom_line(aes(group = short_code), alpha = 0.5)+
  geom_text_repel(data = filter(snotel_predict, snowyrlo_snotel == 0),
                  size = 2.5, segment.color = "black", alpha = 1,
                  segment.size = 0.2, box.padding = 0.1,
                  mapping = aes(label = paste0(short_code, "_", slope))) + 
  xlab("May SWE (cm), snotel") +
  ylab(paste0("Water temperature (", intToUtf8(176), "C), top 2m"))+
  theme_bw() +
  theme(legend.position = "none") +
  theme(strip.text.x = element_text(),
        axis.title.x = element_text(vjust = -0.5)) + 
  scale_colour_viridis_d(end = 0.85) +
  xlim(x = c(0, 325))

ggarrange(plot_wtempSWE_snodas, snodas_predict_plot,
          plot_wtempSWE_snotel, snotel_predict_plot)

```

```{r, include=FALSE}
# Is each predicted ("fit") value equivalent to the actual value?
snodas_predict %>%
  select(site, year, variable, value, fit, SWE_May_snodas) %>%
  {

    round(.$fit, digits = 2) == round(.$value, digits = 2)
    
  } %>%
  # Is the whole vector of logical tests TRUE?
  all()

snotel_predict %>%
  select(site, year, variable, value, fit, SWE_May_snotel) %>%
  {

    round(.$fit, digits = 2) == round(.$value, digits = 2)
    
  } %>%
   # Is the whole vector of logical tests TRUE?
 all()
```

```{r}
# First join coefficient datasets, then update site names, then add
# the static variables
swe_modeling_data <- full_join(x = snodas_models, y = snotel_models,
                               by = c("park", "site"),
                               suffix = c(.x = "_snodas", .y = "_snotel")) %>%
  left_join(x = ., y = match_sites, by = c("site" = "old_name")) %>%
  left_join(x = ., y = physio_subset, by = c("park" = "Park_code",
                                             "site_code"))

```


```{r}
# Set up a tree using all variables, but leave out slope_snodas as a predictor
snodas_tree <- rpart(formula = slope_snodas ~ Elevation_m +
                       Watershed_area_to_lake_area_ratio + Surface_area_ha +
                       Watershed_area_ha + Depth_mean_m + forest + shrub + meadow +
                       barren + snowice,
                     data = swe_modeling_data, minsplit = 10)

plot(x = as.party(snodas_tree),
     main = "SNODAS Slopes",
     terminal_panel = node_boxplot)

plotcp(snodas_tree)

```


```{r}

snotel_tree <- rpart(formula = slope_snotel ~ Elevation_m +
                       Watershed_area_to_lake_area_ratio + Depth_mean_m + forest + 
                       barren,
                     data = swe_modeling_data, minsplit = 10)

plot(x = as.party(snotel_tree),
     main = "SNOTEL Slopes",
     terminal_panel = node_boxplot)

plotcp(snotel_tree)
# If the snotel model is in the ballpark, it’s highest elevation
# forested lakes with shallow depths that have the anomalous slopes

```



### Comparison of weekly (continuous) surface temp vs profile temp @ top 2m
**Note** that weekly temps are not calculated around a specific point, but based on the calendar year. So if we'd like to compare means of weeks centered, e.g., on the date a profile measurement was taken, we'll have to change this. [This](https://www.storybench.org/how-to-calculate-a-rolling-average-in-r/) might show what it would take.
```{r}

daily_temps <- read.csv(file = "../data/analysis_outputs/all-daily-temp-summaries.csv",
                        stringsAsFactors = FALSE) %>%
  select(obs_year:mean_value) %>%
  filter(measure == "SurfaceTemp") %>%
  left_join(x = ., y = match_sites, by = c("lake" = "old_name")) %>%
  select(park, site_code, everything(), -lake)

profile_data <- read_excel(
  path = file.path("..",
                   "data",
                   "NPS_NCCN_Mtn_Lakes_Exports",
                   "qs_b364_Water_Column_Profile_Data_20200723_160058.xlsx"),
  col_types = c("text", "text", "text", "numeric", "date", "logical", "numeric",
                "numeric", "date", "text", "logical", "numeric", "text", "text",
                "text", "date", "text"))%>%
  left_join(x = ., y = match_sites, by = c("Site_name" = "old_name")) %>%
  select(Park_code, site_code, everything(), -Site_code) %>%
  mutate(week_number = week(Start_date))


profile_dates <- profile_data %>%
  select(Park_code, site_code, Start_date, Event_year, week_number)

# Summarize daily temps by week, only for weeks that match to profile temp
# measurement events
weekly_temps <- daily_temps %>%
  mutate(new_date = ymd(truncated = 1,
                        paste(obs_year, obs_month, obs_day, sep = "-")),
         week_number = week(new_date)) %>%
  semi_join(x = ., y = profile_dates, by = c("park" = "Park_code",
                                             "site_code",
                                             "obs_year" = "Event_year",
                                             "week_number")) %>%
  group_by(park, site_code, obs_year, week_number) %>%
  summarise(mean_surf_temp = mean(mean_value))

# Pull the profile temp vals from bigjoin (where they've been aggregated already)
profile_temps <- bigjoin %>%
  select(park_code, site_code, start_date, event_year, variable, value) %>%
  filter(variable == "ProfTemp_top2m") %>%
  mutate(week_number = week(start_date))

# Join daily and profile temps based on week in which they occurred
joined_temps <- full_join(x = weekly_temps, y = profile_temps,
                          by = c("park" = "park_code",
                                 "site_code",
                                 "obs_year" = "event_year",
                                 "week_number")) %>%
  spread(key = variable, value = value) %>%
  # Remove Hoh, because of lake of continuous data (expected, right?)
  filter(site_code != "Hoh")

ggplot(data = joined_temps) +
  geom_point(aes(x = ProfTemp_top2m, y = mean_surf_temp, color = site_code))

ggplot(data = joined_temps) +
  geom_point(aes(x = ProfTemp_top2m, y = mean_surf_temp)) +
  facet_wrap(. ~ site_code) +
  geom_smooth(method = "lm", aes(x = ProfTemp_top2m, y = mean_surf_temp),
              se = FALSE)

```

```{r, include=FALSE}
model_summaries <- map_df(.x = split(x = joined_temps, f = joined_temps$site_code),
                          .f = ~ {
                            
                            temp_lm <- lm(formula = ProfTemp_top2m ~ mean_surf_temp, data = .x)
                            
                            coefs <- temp_lm$coefficients
                            locs <- .x %>% ungroup() %>% select(park, site_code) %>% unique()
                            adj_rsq <- summary(temp_lm)$adj.r.squared
                            coef_p_value <- summary(temp_lm)$coefficients[8]
                            
                            return(data.frame(locs$park, locs$site_code, coefs, adj_rsq, coef_p_value))
                          })
#export R script from Rmd file
#library(knitr)
#purl("analyses_with_static_vars.Rmd")
```















