---
title: "Static Lake Variable Classification"
author: "Matt Brousil"
date: "7/8/2020"
output: html_document
---

## A script that uses PCA and regression tree approaches to classify lakes
## Messed up by Hampton on 10 July 2020
```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(readxl)
library(factoextra)
library(viridis)
library(vegan)
library(rpart)
library(partykit)
library(rpart.plot)
```

```{r}
physio <- read_excel(
  path = file.path("..",
                   "data",
                   "analysis_outputs",
                   "study-site-tables.xlsx"))

# Grab bigjoin for land cover
bigjoin <- readRDS(file = "../data/analysis_outputs/bigjoin.rds")

land_cover <- bigjoin %>% select(park_code, site_code, forest:snowice) %>%
  unique()

# Name matching table
match_sites <- readRDS(file = file.path("..",
                                        "data",
                                        "name_site_matches.rds"))

# Select continuous variables
physio_subset <- physio %>%
  select(Park_code, Lake, Elevation_m, Watershed_area_to_lake_area_ratio,
         Surface_area_ha, Depth_mean_m, Volume_m3,
         Aspect, BlueLineOutlet) %>%
  left_join(x = ., y = match_sites,
            by = c("Lake" = "old_name")) %>%
  select(Park_code, site_code, everything(), -Lake) %>%
  full_join(x = ., y = bigjoin, by = c("Park_code" = "park_code",
                                          "site_code"))

```

Scale a numeric subset of the data
```{r}
physio_scale <- physio_subset %>%
  select(Elevation_m:snowice) %>%
  mutate_if(.predicate = is.numeric, .funs = ~ scale(.)) %>%
  cbind(Park_code = physio_subset$Park_code, .)

row.names(physio_scale) <- paste0(physio$Park_code, "_", physio$site_code)
```


### Run PCA
```{r}
physio_pca <- rda(select_if(physio_scale, .predicate = is.numeric))
```

#### PCA Results:
```{r}
physio_pca
```

```{r, echo=FALSE}
biplot(physio_pca,
       display = c("sites", 
                   "species"),
       type = c("text",
                "points"))
```

Add hulls (currently for park)  
```{r}
# https://rpubs.com/brouwern/veganpca#:~:text=PCA%20(Principal%20Components%20Analysis)%20is,is%20also%20developing%20biplot%20tools).
biplot(physio_pca,
       display = c("sites", 
                   "species"),
       type = c("text",
                "points"))
ordihull(physio_pca,
         group = physio_scale$Park_code)
```


### Run PCA outside of vegan for another plotting option
```{r}
physio_pca <- prcomp(x = select_if(physio_scale,
                                   .predicate = is.numeric), scale. = FALSE)
```

### PCA Results:
```{r}
physio_pca
```

Scree plot
```{r}
fviz_eig(physio_pca)
```

Bioplot & variable contribution
```{r, fig.show="hold", out.width="50%"}


fviz_pca_biplot(X = physio_pca, col.var = viridis(n = 5)[2],
                col.ind = plasma(n = 5)[4], ggtheme = theme_bw())


fviz_pca_var(X = physio_pca, col.var = "contrib", repel = TRUE,
             gradient.cols = viridis(n = 5, begin = 0, end = 1),
             ggtheme = theme_bw())
```


### Regression tree

```{r}
mean_chl <- bigjoin %>%
  select(park_code, site_code, event_year, variable, value) %>%
  filter(variable == "Chlorophyll") %>%
  spread(key = variable, value = value) %>%
  group_by(park_code, site_code) %>%
  summarise(mean_chl = mean(Chlorophyll))
```


```{r}
reg_tree_data <- physio_subset %>%
  full_join(x = ., y = mean_chl, by = c("Park_code" = "park_code",
                                        "site_code"))
```

Specify the tree, setting minimum number of obs per node before a split to be 10
```{r}
tree <- rpart(formula = mean_chl ~ Elevation_m.x +  Watershed_area_to_lake_area_ratio.x + Depth_mean_m.x + forest + barren +
               solar_jas + Volume_m3.x + SWE_May + flush_index_noSWE,
              data = reg_tree_data, minsplit = 10)

plot(x = as.party(tree),
            main = "Chlorophyll",
            terminal_panel = node_boxplot)

plotcp(tree)
ptree <- prune(tree, cp= tree$cptable[which.min(tree$cptable[,"xerror"]),"CP"])
summary(ptree)
rpart.plot(ptree, main="Average Chlorophyll")

```


Based on the cross-validation plot, the ideal tree size is 1...

