---
title: "Sen's regression trees"
author: "Matt Brousil"
date: "2020-04-14"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(rpart)
library(partykit)
```

### Load the necessary data
```{r}
# All variables of interest from NPS
bigjoin <- readRDS(file = "../data/analysis_outputs/bigjoin.rds")

# Sen's slopes, calculated at lake level (i.e., not park-level mean)
sens_data <- read.csv(file = "../data/analysis_outputs/sens_slopes_na_rm_lakelvl.csv",
                      stringsAsFactors = FALSE)
```

#### Look at the data briefly:
```{r}
as_tibble(bigjoin)

as_tibble(sens_data)
```

#### Prepare data for use
Look at the variables from bigjoin that we're interested in. How many of them
have multiple unique values per site? (i.e., multiple years of unique data?)
```{r}
bigjoin %>%
  select(park_code, site_code, year, Elevation_m, elevmean_wshed, Surface_area_ha,
         Watershed_area_ha, Watershed_area_to_lake_area_ratio, Depth_mean_m,
         Depth_max, solar_jas, solar_dec, aspectns_lake, aspectns_wshed, 
         forest, snowice, barren, SWE_May_snotel, salmo_adult_ever, BlueLineInlet,
         BlueLineOutlet) %>%
  unique() %>%
  gather(key = variable, value = value,
         Elevation_m:BlueLineOutlet) %>%
  group_by(site_code, variable) %>%
  distinct(value) %>%
  count(site_code, variable) %>%
  filter(n > 1) %>%
  pull(variable) %>%
  unique()
```

Clean up and join the bigjoin and sens variables:
```{r}
# Summarize variables that aren't static
swe_may_snotel_summarize <- bigjoin %>%
  select(park_code, site_code, year, SWE_May_snotel) %>%
  group_by(park_code, site_code) %>%
  summarize(SWE_May_snotel_mean = mean(SWE_May_snotel, na.rm = T))

# Compile additional bigjoin vars of interest
bigjoin_subset <- bigjoin %>%
  select(park_code, site_code, Elevation_m, elevmean_wshed, Surface_area_ha,
         Watershed_area_ha, Watershed_area_to_lake_area_ratio, Depth_mean_m,
         Depth_max, solar_jas, solar_dec, aspectns_lake, aspectns_wshed, 
         forest, snowice, barren, salmo_adult_ever, BlueLineInlet,
         BlueLineOutlet) %>%
  unique()

bigjoin_subset <- full_join(x = bigjoin_subset, y = swe_may_snotel_summarize,
                            by = c("park_code", "site_code")) 

sens_wide <- sens_data %>%
  spread(key = variable_name, value = slope)

sens_interest <- sens_wide %>%
  select(park, lake,
         # Explanatory sens vars
         AirTemp, SWE_May,
         # Response sens vars
         secchi_value_m, Chlorophyll, SurfTemp, `Total N`, `Total P`, DO_top2m,
         pH_top2m)

regression_data <- full_join(x = bigjoin_subset, y = sens_interest,
                             by = c("park_code" = "park",
                                    "site_code" = "lake"))
```

#### Take a look at variable correlations:
```{r}
corrplot::corrplot(corr = cor(select(regression_data, -park_code,
                                     -site_code, -contains("Blue")),
                              use = "complete"),
                   type = "lower")
```
Look at solar_dec and ice cover correlation
```{r}
ice_data <- bigjoin %>%
  select(park_code, site_code, ice_out_doy, ice_in_doy, snowice) %>%
  unique() %>%
  mutate(ice_length = ice_in_doy - ice_out_doy) %>%
  filter(!is.na(ice_length))

ice_data %>%
  ggplot() +
  geom_point(aes(x = ice_length, y = snowice, color = site_code))
```

Hard to draw much info from this as each lake appears to have a single value for snowice cover. Going to ask NPS what year these data originated from.


#### Regression trees

##### Fit trees for all response vars at once:
```{r}
sens_trees <- map(.x = select(regression_data, secchi_value_m:pH_top2m),
                  .f = ~ rpart(formula = .x ~ AirTemp + SWE_May + Elevation_m +
                                 elevmean_wshed + Surface_area_ha + Watershed_area_ha +
                                 Watershed_area_to_lake_area_ratio + Depth_mean_m +
                                 Depth_max + solar_jas + solar_dec + aspectns_lake +
                                 aspectns_wshed + forest + snowice + barren +
                                 salmo_adult_ever + BlueLineInlet + BlueLineOutlet +
                                 SWE_May_snotel_mean,
                               data = regression_data))
```

##### View trees
```{r}
map2(.x = sens_trees,
     .y = names(sens_trees),
     .f = ~ {
       plot(x = as.party(.x),
            main = paste("Reponse variable: ", .y),
            terminal_panel = node_boxplot)
     })
```


##### Tree diagnostics
```{r}
map2(.x = sens_trees,
     .y = names(sens_trees),
     .f = ~ {
       print(paste("Response variable: ", .y))
       printcp(.x)
       plotcp(.x)
       par(mfrow = c(1,2))
       rsq.rpart(.x)
     })
```